---
title: getting started with bayesian inference
author: ''
date: '2020-06-27'
slug: getting-started-with-bayesian-inference
categories:
  - statistics
tags:
  - statistics
  - R
  - bayesian
description: ''
---


<script type="text/javascript"
  src="https://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML">
</script>




```{r, include=FALSE}

## Function for plotting priors, likelihoods, and posteriors for binomial data
## Output consists of a plot and various statistics
## PS and PF determine the shape of the prior distribution.
## PS = prior success, PF = prior failure for beta dist. 
## PS = 1, PF = 1 corresponds to uniform(0,1) and is default. If left at default, posterior will be equivalent to likelihood
## k = number of observed successes in the data, n = total trials. If left at 0 only plots the prior dist.
## null = Is there a point-null hypothesis? null = NULL leaves it out of plots and calcs
## CI = Is there a relevant X% credibility interval? .95 is recommended and standard



plot.beta <- function(PS = 1, PF = 1, k = 0, n = 0, null = NULL, CI = NULL, ymax = "auto", main = NULL) {
        
        x = seq(.001, .999, .001) ##Set up for creating the distributions
        y1 = dbeta(x, PS, PF) # data for prior curve
        y3 = dbeta(x, PS + k, PF + n - k) # data for posterior curve
        y2 = dbeta(x, 1 + k, 1 + n - k) # data for likelihood curve, plotted as the posterior from a beta(1,1)
        
        if(is.numeric(ymax) == T){ ##you can specify the y-axis maximum
                y.max = ymax
        }        
        else(
                y.max = 1.25 * max(y1,y2,y3,1.6) ##or you can let it auto-select
        )
        
        if(is.character(main) == T){
                Title = main
        }
        else(
                Title = "Prior-to-Posterior Transformation with Binomial Data"
        )
        
        
        plot(x, y1, xlim=c(0,1), ylim=c(0, y.max), type = "l", ylab= "Density", lty = 2,
             xlab= "Probability of success", las=1, main= Title,lwd=3,
             cex.lab=1.5, cex.main=1.5, col = "honeydew4", axes=FALSE)
        
        axis(1, at = seq(0,1,.2)) #adds custom x axis
        axis(2, las=1) # custom y axis
        
        
        
        if(n != 0){
                #if there is new data, plot likelihood and posterior
                lines(x, y2, type = "l", col = "darkslategray3", lwd = 2, lty = 3)
                lines(x, y3, type = "l", col = "firebrick4", lwd = 5)
                legend("topleft", c("Prior", "Posterior", "Likelihood"), col = c("honeydew4", "firebrick4", "darkslategray3"), 
                       lty = c(2,1,3), lwd = c(3,5,2), bty = "n", y.intersp = .55, x.intersp = .1, seg.len=.7)
                
                ## adds null points on prior and posterior curve if null is specified and there is new data
                if(is.numeric(null) == T){
                        ## Adds points on the distributions at the null value if there is one and if there is new data
                        points(null, dbeta(null, PS, PF), pch = 21, bg = "blue", cex = 1.5)
                        points(null, dbeta(null, PS + k, PF + n - k), pch = 21, bg = "darkorchid", cex = 1.5)
                        abline(v=null, lty = 5, lwd = 1, col = "grey73")
                        ##lines(c(null,null),c(0,1.11*max(y1,y3,1.6))) other option for null line
                }
        }
        
        ##Specified CI% but no null? Calc and report only CI
        if(is.numeric(CI) == T && is.numeric(null) == F){
                CI.low <- qbeta((1-CI)/2, PS + k, PF + n - k)
                CI.high <- qbeta(1-(1-CI)/2, PS + k, PF + n - k)
                
                SEQlow<-seq(0, CI.low, .001)
                SEQhigh <- seq(CI.high, 1, .001)
                ##Adds shaded area for x% Posterior CIs
                cord.x <- c(0, SEQlow, CI.low) ##set up for shading
                cord.y <- c(0,dbeta(SEQlow,PS + k, PF + n - k),0) ##set up for shading
                polygon(cord.x,cord.y,col='orchid', lty= 3) ##shade left tail
                cord.xx <- c(CI.high, SEQhigh,1) 
                cord.yy <- c(0,dbeta(SEQhigh,PS + k, PF + n - k), 0)
                polygon(cord.xx,cord.yy,col='orchid', lty=3) ##shade right tail
                
                return( list( "Posterior CI lower" = round(CI.low,3), "Posterior CI upper" = round(CI.high,3)))
        }
        
        ##Specified null but not CI%? Calculate and report BF only 
        if(is.numeric(null) == T && is.numeric(CI) == F){
                null.H0 <- dbeta(null, PS, PF)
                null.H1 <- dbeta(null, PS + k, PF + n - k)
                CI.low <- qbeta((1-CI)/2, PS + k, PF + n - k)
                CI.high <- qbeta(1-(1-CI)/2, PS + k, PF + n - k)
                return( list("BF01 (in favor of H0)" = round(null.H1/null.H0,3), "BF10 (in favor of H1)" = round(null.H0/null.H1,3)
                ))
        }
        
        ##Specified both null and CI%? Calculate and report both
        if(is.numeric(null) == T && is.numeric(CI) == T){
                null.H0 <- dbeta(null, PS, PF)
                null.H1 <- dbeta(null, PS + k, PF + n - k)
                CI.low <- qbeta((1-CI)/2, PS + k, PF + n - k)
                CI.high <- qbeta(1-(1-CI)/2, PS + k, PF + n - k)
                
                SEQlow<-seq(0, CI.low, .001)
                SEQhigh <- seq(CI.high, 1, .001)
                ##Adds shaded area for x% Posterior CIs
                cord.x <- c(0, SEQlow, CI.low) ##set up for shading
                cord.y <- c(0,dbeta(SEQlow,PS + k, PF + n - k),0) ##set up for shading
                polygon(cord.x,cord.y,col='orchid', lty= 3) ##shade left tail
                cord.xx <- c(CI.high, SEQhigh,1) 
                cord.yy <- c(0,dbeta(SEQhigh,PS + k, PF + n - k), 0)
                polygon(cord.xx,cord.yy,col='orchid', lty=3) ##shade right tail
                
                return( list("BF01 (in favor of H0)" = round(null.H1/null.H0,3), "BF10 (in favor of H1)" = round(null.H0/null.H1,3),
                             "Posterior CI lower" = round(CI.low,3), "Posterior CI upper" = round(CI.high,3)))
        }
        
}


```



[In my previous post](https://apoorvasrinivasanblog.com/post/hypothesis-testing/), we spoke about hypothesis testing from a frequentist perspective. If you took a stats class, that is the method you probably learn. There is another, arguably better method to conduct a hypothesis testing and that using bayesian inference.


My goal in this blogpost is to break down the basics of bayesian inference.

This blogpost assumes prior knowledge on conditional, marginal and random variable concepts.


### why go bayes?

there's tons of material on this by poeple 

#### bayes theorem

Before getting into the inference, let's take a moment to talk about bayes theorem. As one of the most popular formula in statistics, it needs no introduction but it may be helpful to state it here again.

Mathematical definition of bayes: $P(A/B) = \dfrac {P(B/A)P(A)}{P(B)}$



There are three main ingredients to bayesian hypothesis:
1) Prior
2) Likelihood
3) Posterior

They can be put together into the bayes theorem as shown:

$posterior = \dfrac {prior * likelihood} {marginal \hspace{0.2cm} likelihood}$


Let's start with an example:


Let's say you have a headache and sore throat, and you know that some people in the office you work at have also had the flu. Does this mean you have the flu? Let's find out if bayesian inference can help us figure it out.

Putting our problem in mathematical terms, it looks like this:

$P(flu| symptoms) = \dfrac {P(symptoms| flu) * P(flu)} {P(symptoms)}$ 

# prior

The first ingredient is the background knowledge on the parameters of the model being tested. This refers to the knowledge available _before_ seeing the data. 


Let's assume you don't know anything about the prevalence of flu. So you'll probably pick a prior where all probabilities are equally likely.  It will look something like this

<plot the prior plot>

```{r}
plot.beta(1,1,ymax=3.2,main="Uniform Prior, Beta(1,1)")
```


Wanting to gain a little more information you roll over, grab your phone
and search Google. You find an article that says that only 12% of the
population will get the flu in a given year. But another article says that the prevalence could be a little higher. In this case a distribution of our prior belief might be more appropriate to incorporate our uncertainity. This distribution is known as the __prior distribution__. 

<plot another prior dist>


```{r}
plot.beta(12,88,ymax=15,main="Informed Prior, Beta(12, 88)")
```


:pencil:
In Bayesian inference, the beta distribution is the conjugate prior probability distribution for the binomial distribution. Computing a posterior using a conjugate prior is very convenient, because you can avoid expensive numerical computation. You can learn more about conjugate priors here(add link)


# likelihood

The second ingredient is likelihood, this is where we extract information from our data. Likelihood function is the one that does all the work in bayesian inference. It is also one of the trickier concepts to understand in one shot.

Lets say you go out and collect data from 100 of your colleagues and 30 of them are showing symptoms. The likelihood curve wil look like this:


```{r}
x = seq(.001, .999, .001) ##Set up for creating the distributions
y2 = dbeta(x, 1 + 30, 1 + 70) # data for likelihood curve, plotted as the posterior from a beta(1,1)

plot(x, y2, xlim=c(0,1), ylim=c(0, 1.25 * max(y2,1.6)), type = "l", ylab= "Density", lty = 3,
     xlab= "Probability of success", las=1, main="Likelihood Curve for Flu Data",lwd=2,
     cex.lab=1.5, cex.main=1.5, col = "cyan3", axes=FALSE)
axis(1, at = seq(0,1,.2)) #adds custom x axis
axis(2, las=1) # custom y axis


```


The Likelihood Principle states that the likelihood function contains all of the information relevant to the evaluation of statistical evidence. In other words, facets of the data that do not factor into the likelihood function are irrelevant to the evaluation of the strength of the statistical evidence. Likelihood distribution is the curve that best fits the data we're given.



# is likelihood a probability?

Likelihood is not a probability, but it's proportional to it. 

The likelihood of a hypothesis (H) given some data (D) is proportional to the probability of obtaining D given that H is true, multiplied by an arbitrary positive constant (K).

Since a likelihood isn’t actually a probability it doesn’t obey various rules of probability. For example, likelihood need not sum to 1.

A critical difference between probability and likelihood is in the interpretation of what is fixed and what can vary. The distinction is subtle, so pay close attention. For the conditional probability, the hypotheses is fixed and the data vary. For likelihood, the data are a given and the hypotheses vary. 



# posterior

Now for the easiest part. Getting the posterior distribution. This is the distribution representing our belief about the parameter values after we have calculated everything on the right hand side taking the observed data into account.


$Posterior \propto Prior * Likelihood$

### why did we ignore the demonimator?

Well, apart from being the marginal distribution of the data it doesn’t really have a fancy name, although it’s sometimes referred to as the evidence. Remember, we’re only interested in the parameter values but P(data) doesn’t have any reference to them. In fact, P(data) doesn’t even evaluate to a distribution. It’s just a number. we’ve already observed the data so we can calculate P(data). In general, it turns out that calculating P(data) is very hard and so many methods exist to calculate it. 

The reason why P(data) is useful is because the number that comes out is a normalising constant. One of the necessary conditions for a probability distribution is that the sum of all possible outcomes of an event is equal to 1. The normalising constant makes sure that the resulting posterior distribution is a true probability distribution by ensuring that the sum of the distribution (I should really say integral because it’s usually a continuous distribution) is equal to 1.


Coming back to our example, so far we have two distributions: prior and likelihood. We can get the posterior by multiplying the two. The resulting distribution is shown below

```{r}
#Initial Priors and final Posteriors after all rounds at once

#par(mfrow=c(1,2))
plot.beta(1,1,30,100,ymax=10,main="Beta(1,1) to Beta(59,43)")

```

With the uniform prior, the posterior falls right on top of the likelihood function. In the plot above, the likelihood is nearly invisible because the posterior sits right on top of it. When the prior has only 1 or 2 data points worth of information, it has essentially no impact on the posterior shape 


```{r}
plot.beta(12,88,30,100,ymax=15,main="Beta(12,88) to Beta(30,70)")
```

The second plot above shows how the posterior splits the difference between the likelihood and the informed prior .My takeaway is that the the probability of  having the flu based on the data you've collected is about 20% which is slightly higher than the  prior.


